Towards planning of infantile cardiac surgery based on simulation

### 2012-02-29 Doc/Miccai/shell-model.tex
modifs in section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@348 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
To do a planning of the surgery, the geometry changes of these vessel walls (the motions and deformations, their configuration when they are cut or stitched)  must be anticipated.

@@

Several different models are commonly used for modelling thin structures or
tubular-like structures resembling blood vessels. \tomas{Add refs. or
reformulate!} For the predictive simulation of the surgery we choose to use
thin shell elements, because in our opinion it satisfies all the
requirements very well. They are very often considered numerically too
expensive, this is however not necessarily true.

\tomas{image: single shell}

The formulation of the thin shell elements is based on the mechanics of
continuum. These elements are commonly used in physics and mechanics for
deformation modelling of thin structures because they are numerically more
stable then volumetric FEM methods in this situation. The shell element is
a 2-dimensional element whose thickness is a parameter of the simulation.
Often the thickness is considered constant over the surface of the element.
Shell combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
$\myvec{x}$ from the rest shape $\myvec{x_0}$.

\begin{equation}
  \myvec{f} = K(\myvec{x} - \myvec{x_0})
 \label{linear_static_model} 
\end{equation}

\christian{I think that we should precise that we rely on corotational shell formulation: the model is linear in the local frame of the element, otherwise equation (\ref{linear_static_model}) is false } 
For more thorough description we refer the reader to the textbook about
Finite Element methods for example \cite{Reddy1993}.

In the rest of the section we want to show that the performance of a model
based on shell elements is adequate for the simulation.

@@

Figure \ref{fig-deformations} shows different deformations that are likely to
occur during the surgery (bending and twisting). The figure compares our
results with the results of Li et al. \cite{Li2009}. As you can see the
results are closer to the real deformations.

@@

We have used the model of Olivier Comas \cite{Comas2010b,Comas2010c} and
also our own model. But our method described in section \ref{sec-method}
does not depend on any particular model. Since we use the method as a
geometrical tool we have also certain independence on physical parameters.
For example the variation of Young's modulus in the interval from $10^3$ to
$10^7$ resulted to only $0.1\%$ change of the deformation.

>>>>

\subsubsection{Requirements: } 
Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
To do a planning of the surgery, the geometry changes of these vessel walls (the motions and deformations, their configuration when they are cut or stitched)  must be anticipated.

@@

In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
The options on the choice of model are limited: 
The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline? }  ) must be discarded because they do not support cut of the wall along the vessel axis. 
Mass-springs approaches that are often used for cloth simulation (\christian{do we consider this option ? if yes add ref ??}), allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
Moreover, we show that we can still obtain fast computation with a reasonable number of elements.

\subsubsection{Thin shell elements: } 
The formulation of the thin shell elements is based on the mechanics of continuum. 
\tomas{image: single shell}
For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
The shell element is a 2-dimensional element (in our case, triangle element) whose thickness is a parameter of the simulation.
Often the thickness is considered constant over the surface of the element.
Shell combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
$\myvec{x}$ from the rest shape $\myvec{x_0}$.
To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ?? }).

In this local frame, the computation of the internal forces is linear:
\begin{equation}
  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
 \label{linear_static_model} 
\end{equation}
where $\mathbf{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}. 
We emphasis that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used).
%
When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
\begin{equation}
\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
\end{equation}
Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.

\subsubsection{Accuracy: } 
In the following of the section we want to show that the performance of a model
based on shell elements is adequate for the simulation.

@@

Figure \ref{fig-deformations} shows different deformations that are likely to
occur during the surgery (bending and twisting). The figure compares our
results with the results of Li et al. \cite{Li2009}. As you can see the
results are close to the real deformations.

@@

\christian{ I think this should be put in the result section:
Since we use the method as a
geometrical tool we have also certain independence on physical parameters.
For example the variation of Young's modulus in the interval from $10^3$ to
$10^7$ resulted to only $0.1\%$ change of the deformation.
}
